---
title: "Assignment 2: Coding Basics"
author: "Meghan Seyler"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
increasingbyfours<-seq(1,100,4) #here I created a sequence that increases by fours and assigned it the name increasingbyfours
increasingbyfours 

#2. 
mean_increasingbyfours<-mean(increasingbyfours) #taking the mean of increasingbyfours and assigning the value a name
mean_increasingbyfours
median_increasingbyfours<-median(increasingbyfours) #taking the median of increasingbyfours and assigning the value a name
median_increasingbyfours

#3. 
#testing statements that will generate a true or false output. The mean is NOT greater than the median
mean_increasingbyfours>median_increasingbyfours  
mean_increasingbyfours<median_increasingbyfours
mean_increasingbyfours==median_increasingbyfours


```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5. 
studentnames<-c ("Aislinn", "Tay", "Eva", "John") #character

testscores<-c (96,60,45,5) #double

pass<-c (TRUE,FALSE,FALSE,FALSE) #logical

#6. 
typeof(studentnames)
typeof(testscores)
typeof(pass)

#7 
studentscores<- data.frame(studentnames,testscores,pass)
studentscores

#8
colnames(studentscores)<- c("First_Name", "Total_Score", "PASS(T/F)")
studentscores
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: In a matrix all columns must have the same mode and length. A data frame can consist of many modes. This is demonstrated by the data frame above which consists of logical, character, and numerical modes.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
determinepass<- function (testscores) {
  if(x>=50) 
    x==TRUE
  else {x==FALSE}
}
determinepass(49)
```

```{r}
determinepass2<- ifelse(testscores>=50, "Pass", "Fail")
determinepass2
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
