---
title: "Assignment 09: Data Scraping"
author: "Meghan Seyler"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()

library(tidyverse)
library(rvest)
library(lubridate)

mytheme_A09 <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme_A09)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
LWSP_Durham_URL <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020'

LWSP_Durham_webpage <- read_html(LWSP_Durham_URL)
LWSP_Durham_webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- LWSP_Durham_webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name

pswid <- LWSP_Durham_webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pswid

ownership <- LWSP_Durham_webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max.withdrawals.mgd <- LWSP_Durham_webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
#put scraped data into a dataframe and add date column
LWSP_Durham_df<- data.frame(
  "Month" = rep(1:12),
  "Year" = rep(2020,12),
  "Water_System_Name" = water.system.name, 
  "PSWID" = pswid,
  "Ownership" = ownership,
  "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)
)

LWSP_Durham_df

#Modify the dataframe to include the facility name and type as well as the date (as date object)
LWSP_Durham_df <- LWSP_Durham_df %>% 
  mutate(Date = my(paste(Month,"-",Year)))

LWSP_Durham_df

#5
max_withdrawls_plot<-ggplot(LWSP_Durham_df,aes(x=Date,y=Max_Withdrawals_mgd)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2020 Water usage data for",water.system.name),
       subtitle = paste("PSWID", pswid),
       y="Withdrawal (mgd)",
       x="Date")
max_withdrawls_plot

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.
#the base URL
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
the_pswid <- '03-32-010'
the_year <- 2020
the_scrape_url <- paste0(the_base_url, '/', the_pswid, '/', the_year)
print(the_scrape_url)

scrape.it <- function(the_year, pswid_tag){
  
  #Retrieve the website contents 
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?', 
                                  pswid_tag, '/', the_year))
  
  #Set the element address variables
    water.system.name_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
    pswid_tag <- 'td tr:nth-child(1) td:nth-child(5)'
    ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
    max.withdrawals.mgd_tag <- 'th~ td+ td'
  
  #Scrape the data items
    water.system.name <- LWSP_Durham_webpagee %>% html_nodes(water.system.name_tag) %>% html_text()
    pswid <- LWSP_Durham_webpage %>%   html_nodes(pswid_tag) %>%  html_text()
    ownership <- LWSP_Durham_webpage %>% html_nodes(ownership_tag) %>% html_text()
    max.withdrawals.mgd <- LWSP_Durham_webpage %>% html_nodes(max.withdrawals.mgd_tag) %>% html_text()
  
  #Convert to a dataframe
  df_withdrawals_max <- data.frame("Month" = rep(1:12),
                               "Year" = rep(the_year,12),
                               "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>% 
    mutate(Water_System_Name = !!water.system.name,
           pswid = !!pswid,
           ownership = !!ownership,
           Date = my(paste(Month,"-",Year)))
  
  #Pause for a moment - scraping etiquette
  #Sys.sleep(1) #uncomment this if you are doing bulk scraping!
  
  #Return the dataframe
  return(df_withdrawals_max)
}
```


7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
#Run the function
the_df <- scrape.it(2015,'0004-0001')
view(the_df)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Durham.2015.data}
#8


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9


```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?